Pangenome graph layout by Path-Guided Stochastic Gradient Descent

### 2023-08-07 manuscript/fig_2d_layouts.tex
incorporate feedback from joerg
fix compilation issue

\begin{figure*}[!htb]
	\centering
	\begin{subfigure}{1.0\textwidth}
		\centering
		\caption{}
		\includegraphics[width=1.0\linewidth]{fig/2D/hprc-v1.0-pggb.og.lay.H4000.r4.0-highlighted.png}
		\label{fig:sfig1}
	\end{subfigure}
	\\
	\begin{subfigure}{1.0\textwidth}
		\centering
		\caption{}
		\includegraphics[width=1.0\linewidth]{fig/2D/grch38.chr6.MHC_annotated.png}
		\label{fig:sfig2}
	\end{subfigure}
	\\
\begin{subfigure}{1.0\textwidth}
	\centering
	\caption{}
	\includegraphics[width=1.0\linewidth]{fig/2D/grch38.chr6.MHC_genes_annotated.png}
	\label{fig:sfig3}
\end{subfigure}
	\\
\begin{subfigure}{1.0\textwidth}
	\centering
	\caption{}
	\includegraphics[width=1.0\linewidth]{fig/2D/grch38.chr6.C4_gene_annotated.png}
	\label{fig:sfig4}
\end{subfigure}	
	\caption{2D visualizations of all chromosomes of the Human Pangenome Resource Consortium (HPRC) 90 haplotypes pangenome graph, chromosome 6, the major histocompatibility complex (MHC), and the complement component 4 (C4). \textbf{(a)} \textit{odgi draw} layout of the HPRC pangenome graph 90 haplotypes. Displayed are all 24 autosomes and the mitochondrial chromosome. \textbf{(b)} \textit{gfaestus} screenshot of the chromosome 6 layout. Colored in blue is the MHC. \textbf{(c)} \textit{gfaestus} screenshot of the MHC. All MHC genes are color annotated. \textbf{(d)} \textit{gfaestus} screenshot of the region around C4, specifically color highlighting genes C4A and C4B.}
	\label{fig:2d_layouts}
\end{figure*}

>>>>

\begin{figure*}[!htb]
	\centering
	\begin{subfigure}{1.0\textwidth}
		\centering
		\caption{}
		\includegraphics[width=1.0\linewidth]{fig/2D/hprc-v1.0-pggb.og.lay.H4000.r4.0-highlighted.png}
		\label{fig:sfig1}
	\end{subfigure}
	\\
	\begin{subfigure}{1.0\textwidth}
		\centering
		\caption{}
		\includegraphics[width=1.0\linewidth]{fig/2D/grch38.chr6.MHC_annotated.png}
		\label{fig:sfig2}
	\end{subfigure}
	\\
	\begin{subfigure}{1.0\textwidth}
		\centering
		\caption{}
		\includegraphics[width=1.0\linewidth]{fig/2D/grch38.chr6.MHC_genes_annotated.png}
		\label{fig:sfig3}
	\end{subfigure}
	\\
	\begin{subfigure}{1.0\textwidth}
		\centering
		\caption{}
		\includegraphics[width=1.0\linewidth]{fig/2D/grch38.chr6.C4_gene_annotated.png}
		\label{fig:sfig4}
	\end{subfigure}	
	\caption{2D visualizations of all chromosomes of the Human Pangenome Resource Consortium (HPRC) 90 haplotypes pangenome graph, chromosome 6, the major histocompatibility complex (MHC), and the complement component 4 (C4). 
		\textbf{(a)} \textit{odgi draw} layout of the HPRC pangenome graph 90 haplotypes. 
		Displayed are all 24 autosomes and the mitochondrial chromosome.
		A red rectangle highlights chromosome 6 which is shown in the subfigure below. 
		\textbf{(b)} \textit{gfaestus} screenshot of the chromosome 6 layout. 
		Colored in blue is the MHC. 
		The hairball in the middle is the centromere. 
		The black structures in the centromere are edges. 
		\textbf{(c)} \textit{gfaestus} screenshot of the MHC. 
		All MHC genes are color annotated and the names of the genes appear as a text overlay. 
		\textbf{(d)} \textit{gfaestus} screenshot of the region around C4, specifically color highlighting genes C4A and C4B. 
		The black lines are the edges of the graph.}
	\label{fig:2d_layouts}
\end{figure*}

### 2023-08-07 manuscript/main.tex
fix compilation issue

	PG-SGD's pseudocode is shown in Algorithm~\ref{alg:pg_sgd} and key steps are sketched in Figure \FIXME{Add puritan figure.}.
	In brief, the algorithm moves one pair of nodes $( v_i, v_j )$ at a time, minimizing the difference between the layout distance $ld_{ij}$ of the two nodes and the nucleotide distance $nd_{ij}$ of the same nodes as calculated along a random path that traverses them.
	%Our multithreaded implementation (\url{https://odgi.readthedocs.io/en/latest/rst/tutorials/sort_layout.html}) presents a working prototype that is based on succinct graph data structures (\url{https://doi.org/ghqdzw}):
	$v_i$ is the node associated with the step $s_i$ sampled uniformly from all the steps in $\mathcal{P}$.
	$v_j$ is the node associated with the step $s_j$ sampled from the same path of $s_i$ by drawing a Zipfian distribution \citep{Zipf1932}.
	The difference between $nd_{ij}$ and $ld_{ij}$ guides the update of the node coordinates in the layout.
	The magnitude $r$ of the update depends on the learning rate $\mu$ of the SGD.
	The number of iterations steers the annealing step size $\eta$ which determines the learning rate $\mu$.
	%A key property of pangenome graph layouts is that their nodes can be arranged so that most edges go between nodes that are close together in the node order \citep{Guarracino2022}.
	A large $\eta$ in the first iterations leads to a globally linear (in 1D) or planar (in 2D) layout.
	By decreasing $\eta$, the layout adjustments become more localized, ensuring that the nodes are positioned in a way to best reflect the nucleotide distances in the paths (i.e., in the genomes). % LIAR: down to the single base pair level.

@@

@@

	%We are grateful to members of the HGSVC and HPRC production teams for their development of resources used in our exposition.
	%We thank the authors of the pangenome resources made available on GenBank which have made our experiments possible.
	We thank Matthias Seybold from the Quantitative Biology Center for maintaining the Core Facility Cluster.

>>>>

	PG-SGD's pseudocode is shown in Algorithm~\ref{alg:pg_sgd} and 2D key steps are sketched in Figure~\ref{fig:sketches}.
	In brief, the algorithm moves one pair of nodes $( v_i, v_j )$ at a time, minimizing the difference between the layout distance $ld_{ij}$ of the two nodes and the nucleotide distance $nd_{ij}$ of the same nodes as calculated along a path that traverses them.

	%Our multithreaded implementation (\url{https://odgi.readthedocs.io/en/latest/rst/tutorials/sort_layout.html}) presents a working prototype that is based on succinct graph data structures (\url{https://doi.org/ghqdzw}):
	$v_i$ is the node associated with the step $s_i$ sampled uniformly from all the steps in $\mathcal{P}$.
	$v_j$ is the node associated with the step $s_j$ sampled from the same path of $s_i$ by drawing a Zipfian distribution \citep{Zipf1932}.
	The difference between $nd_{ij}$ and $ld_{ij}$ guides the update of the node coordinates in the layout.
	The magnitude $r$ of the update depends on the learning rate $\mu$ of the SGD.
	The number of iterations steers the annealing step size $\eta$ which determines the learning rate $\mu$.
	%A key property of pangenome graph layouts is that their nodes can be arranged so that most edges go between nodes that are close together in the node order \citep{Guarracino2022}.
	A large $\eta$ in the first iterations leads to a globally linear (in 1D) or planar (in 2D) layout.
	By decreasing $\eta$, the layout adjustments become more localized, ensuring that the nodes are positioned in a way to best reflect the nucleotide distances in the paths (i.e., in the genomes). % LIAR: down to the single base pair level.

@@

	\FIXME{In the section above we need to clarify that we move the ends of the nodes in 2D, and just the nodes in 1D.}

	\input{fig_sketches}

@@

	%We are grateful to members of the HGSVC and HPRC production teams for their development of resources used in our exposition.
	%We thank the authors of the pangenome resources made available xon GenBank which have made our experiments possible.
	We thank Matthias Seybold from the Quantitative Biology Center for maintaining the Core Facility Cluster.
